Removing nonlinear batch effects using adversarial deep neural networks

### 2019-03-05 thesis.tex
Tidy up the background

\section{Background}

\subsection{Hook / funnel}

\subsection{Awesome stuff that's been done, noble failure}

\subsection{To address this question, we need...}

\subsection{In this study, we address this question by addressing the following hypotheses}

When working with biological data, problems can arise from combining data from multiple sources.
For example, if two labs quantify the sequences of the same RNA samples for the same lung cancer patients but use different equipment to do so, each lab will come up with slightly different measurements.
If a third party were to then study the patients using data from both labs, the differences in measurements could alter their analysis.
These systemic effects are known as ``batch effects'' and are understood to have a nontrivial impact on high-throughput biological data \citep{leek_tackling_2010}.
In one study, researchers found that, contrary to previous knowledge, molecular data from mice and humans clustered more closely by species than by tissue type \citep{yue_comparative_2014}.
Referees showed in a rebuttal that when accounting for batch effects, these data actually clustered more closely by tissue type, as initially expected \citep{gilad_reanalysis_2015}.
Systemic bias can be even more pronounced when using data collected from different tissue types and can skew sec:results, such as in pan-cancer analyses \citep{dayton_classifying_2017-1}.

Several methods exist for removing batch effects from biological datasets.
Two commonly used methods are ComBat \citep{johnson_adjusting_2007} and SVA \citep{leek_capturing_2007}.
ComBat uses an empirical Bayes method to estimate batch effect parameters and then uses linear regression to remove the effects, and SVA uses singular value decomposition to model batch effects which can then be accounted for in statistical analyses.
Since both of these methods use linear adjustments to remove batch effects, they do not account for nonlinear effects.
However, recent advances in neural networks have introduced new ways to account for higher-order, nonlinear relationships in data.

Artificial neural networks are a machine learning tool inspired by the way human brains function; input values pass through layers of linear and nonlinear functions, the final output values are measured against objectives, and the layers of functions are adjusted to bring the outputs closer to the objectives.
This process is repeated until the outputs are sufficiently close to their targets \citep{schmidhuber_deep_2015}.
Autoencoders are a type of neural network that encode and then reconstruct their input, and their traditional objective function is to construct the output as similarly to the input as possible \citep{hinton_reducing_2006}.
Neural networks have historically decreased in effectiveness when working with data from multiple domains \citep{ganin_domain-adversarial_2015}, in part because they may learn based on confounding effects (e.g. which researcher collected the data) instead of learning based on biologically interesting causal effects (e.g. which gene is consistently upregulated in a disease) \citep{louizos_causal_2017-2}.
Recently, researchers have experimented with discouraging the networks from distinguishing between classes by giving them two objective functions:
1. to learn as much as possible about the input data and
2. to forget any patterns that help distinguish between classes \citep{ganin_domain-adversarial_2015,tzeng_deep_2014-2}.
This type of dual objective function has been used in conjunction with autoencoders \citep{louizos_variational_2015}; the autoencoder is ``rewarded'' for reconstructing the input faithfully, but ``punished'' for keeping class-specific patterns.
Research has also shown that neural networks are effective in working with biological data; for example, neural networks have been applied to extract biologically relevant latent spaces in RNA-Seq data with as few as 10,000 samples \citep{way_extracting_2017}, and some researchers have used that latent space to generate realistic synthetic biomedical data for other scientific studies \citep{beaulieu-jones_privacy-preserving_2017}.

Our approach will build on this previous research by creating an adversarial autoencoder that removes confounding effects from multi-origin data.
This autoencoder will be packaged and made available to allow other researchers to remove these effects from their own data to enable further analysis.

@@

\noindent\rule{\textwidth}{1pt}

Many tools have been developed to account and adjust for batch effects in expression data.
However, most of these tools only adjust for linear batch effects without accounting for nonlinear interactions between variables.
Because of the large number of features in expression data and the nature of these features (i.e. transcripts and proteins that interact biologically), we anticipate that these nonlinear effects may still be present even after adjustment with popular tools.
A few notable examples \cite{shaham_removal_2017,shaham_batch_2018} use deep neural networks to correct for batch effects, but these are optimized for use in large, balanced, dual-batched datasets and are therefore unusable for many real-world examples.

\textbf{Hypothesis:}
Using an adversarial neural network can correct for batch effects more completely than previous tools do. % clarify here what "more completely" means

@@

\section{Discussion}

\section{Conclusions}

>>>>

\section{Background} \label{sec:background}

When working with biological data, problems can arise from combining data from multiple sources.
For example, if two labs quantify the sequences of the same RNA samples for the same lung cancer patients but use different equipment to do so, each lab will come up with slightly different measurements (see Figure \ref{fig:workflow}).
If a third party were to then study the patients using data from both labs, the differences in measurements could alter their analysis.
These systemic effects are known as ``batch effects'' and are understood to have a nontrivial impact on high-throughput biological data \citep{leek_tackling_2010}.
In one study, researchers found that, contrary to previous knowledge, molecular data from mice and humans clustered more closely by species than by tissue type \citep{yue_comparative_2014};
however, referees showed in a rebuttal that when accounting for batch effects, these data actually clustered more closely by tissue type, as initially expected \citep{gilad_reanalysis_2015}.
Systemic bias can be even more pronounced when using data collected from different tissue types and can skew sec:results, such as in pan-cancer analyses \citep{dayton_classifying_2017-1}.

@@

Several methods exist for removing batch effects from biological datasets.
Two commonly used methods are ComBat \citep{johnson_adjusting_2007} and SVA \citep{leek_capturing_2007}.
ComBat uses an empirical Bayes method to estimate batch effect parameters and then uses linear regression to remove the effects, and SVA uses singular value decomposition to model batch effects which can then be accounted for in statistical analyses.
Since both of these methods use linear adjustments to remove batch effects, they do not account for nonlinear effects.
As machine learning becomes more common in biological research, these nonlinear batch effects become more troublesome since many machine learning algorithms can successfully identify complex interactions between variables.
Advances in neural networks have introduced new ways to account for higher-order, nonlinear relationships in data.
These networks have proven effective in removing irrelevant, domain-specific signal in credit rating, online reviews, and image recognition tasks \citep{louizos_variational_2015}.
Recent work \cite{shaham_removal_2017,shaham_batch_2018,upadhyay_removal_2019} has applied neural networks to batch effects.
However these recent programs have several issues that make them unusable on many real-world datasets:
they require that the input data only contains two batches, that the batches are perfectly balanced, and that the batches are sufficiently large.
These requirements rarely hold in existing datasets; for example, the bladderbatch dataset used in the documentation for the R SVA package \cite{leek_bladderbatch_2017,leek_sva_2017} has 5 batches, only 57 samples, and between 4 and 19 samples per batch.
Additionally, the statistics these studies have used to validate results don't test whether complex interactions still remain in the data.

To address these issues, we implemented a deep neural network approach to batch correction that works on real-world datasets, and we calculated batch removal measurements based on machine learning algorithms to test for remaining nonlinear effects.
Artificial neural networks are a machine learning tool inspired by the way human brains function; input values pass through layers of linear and nonlinear functions, the final output values are measured against objectives, and the layers of functions are adjusted to bring the outputs closer to the objectives.
This process is repeated until the outputs are sufficiently close to their targets \citep{schmidhuber_deep_2015}.
Autoencoders are a type of neural network that encode and then reconstruct their input, and their traditional objective function is to construct the output as similarly to the input as possible \citep{hinton_reducing_2006}.
Neural networks have historically decreased in effectiveness when working with data from multiple domains \citep{ganin_domain-adversarial_2015}, in part because they may learn based on confounding effects (e.g. which researcher collected the data) instead of learning based on biologically interesting causal effects (e.g. which gene is consistently upregulated in a disease) \citep{louizos_causal_2017}.
Recently, researchers have experimented with discouraging the networks from distinguishing between classes by giving them two objective functions:
1. to learn as much as possible about the input data and
2. to forget any patterns that help distinguish between classes \citep{ganin_domain-adversarial_2015,tzeng_deep_2014}.
This type of dual objective function has been used in conjunction with autoencoders \citep{louizos_variational_2015} where the autoencoder is ``rewarded'' for reconstructing the input faithfully, but ``punished'' for keeping class-specific patterns.
Research has also shown that neural networks are effective in working with biological data; for example, neural networks have been applied to extract biologically relevant latent spaces in RNA-Seq data with as few as 10,000 samples \citep{way_extracting_2017}, and some researchers have used that latent space to generate realistic synthetic biomedical data for other scientific studies \citep{beaulieu-jones_privacy-preserving_2017}.

In this study, we present Confounded, an adversarial autoencoder that identifies and removes effects between different batches.
We test the hypothesis that using an adversarial neural network can correct for batch effects more completely than previous tools do.
We also explore the extent to which batch effects still remain in different datasets after adjustment with various algorithms, and we present a framework to assess how much nonlinear effects between batches still remains after adjustment using various classification algorithms.

@@

\section{Discussion} \label{sec:discussion}

Most exciting takeaways

\begin{enumerate}
	\item Nonlinear interactions still remain after ComBat for larger datasets
	\item Simple metrics aren't enough to identify the nonlinear interactions, but different classifiers help
	\item ComBat works really well for small, traditional datasets
	\item Confounded removes nonlinear effects in larger datasets
\end{enumerate}

\section{Conclusions} \label{sec:conclusion}
